A Possibilistic Valid-time Model

### 2012-04-09 ipmu2012/proposal.tex
last version
I think this should be the final version... but how knows..

In this section, a valid-time model, able to handle imperfect valid-time notions, will be presented. First, the representation and storage of valid-time will be presented. Next, an approach to query the model is proposed. Finally, an example is given.
%The proposal consist on a possibilistic valid-time model. The representation and the querying are explained in the following subsections.

@@

As mentioned in \cite{Pon11}, this approach differs from the one where a valid-time period is represented by one fuzzy set. Such a fuzzy set is seen as a possibility distribution on $\mathbb{R}$ and thus defines just one ill-known value. However, in the presented approach, a time period is modelled using an ill-known set, which is defined by a possibility distribution on $\Pow(\mathbb{R})$.

%The interval has a starting and an ending points. An ill-known valid-time interval is an interval in witch one or both points are ill-known.

%\begin{definition}
%A Possibilistic Valid-Time Period \textbf{PVP} is a possibilistic interval defined by means of two ill-known points, namely $\left[ X,\ Y \right]$
%\begin{equation}
%PVP = \left[X,\ Y \right] 
%\end{equation}
%$X$ and $Y$ are ill-known values in the set of the real numbers $\mathbb{R}$. The uncertainty about the values taken by $X$ and $Y$ are given by the possibility distributions $\pi_X$ and $\pi_Y$.
%\end{definition}

%For convenience, the possibility distributions $\pi_X$ and $\pi_Y$ are given in the way of a triangular distribution, as explained in subsection \ref{subsec:fuzzy-numbers}. This representation allows overlapping (Fig. \ref{fig:pvp}). Note that while the two ill-known values $X,Y \in \mathbb{R}$, the fuzzy interval  $[X,Y] \in \mathbb{R}^2$.

%\begin{figure}[h!]
%  \centering
%  \input{./graphs/2_triangular.tex}
%  \caption{Two fuzzy numbers $X$ and $Y$ denoting a Possibilistic Valid-Time Period \emph{PVP}.}
%  \label{fig:pvp}
%\end{figure}

\subsubsection{Storage}
\label{subsec:storage}
To store a PVP, the possibility distributions defining the ill-known start and end point are stored. To store such a possibility distribution, the representation as used in the fuzzy interface for relational databases \emph{FIRST}~\cite{Medina94gefred.a}, \cite{Gal98} is used. Using this representation, it is possible to represent not only fuzzy numbers, but also (fuzzy) constants. To store an ill-known value, four values (FT, F1, F2 and F3) are stored. They are explained in Table \ref{table:relational-representation-pvp}. Note that while NULL denotes the fuzzy constant NULL, N denotes a null value in the database \cite{Medina94gefred.a}.

%\begin{itemize}
%\item
%\emph{NULL}: This constant refers to a completely ignorance about the value. The possibility distribution for a given fuzzy number $X$ is not defined, therefore, any comparison between a fuzzy number and the \emph{NULL} constant always returns $0$.
%\item
%\emph{UNKNOWN}: % The possibility distribution for a given fuzzy number $X$ is $\pi_X=1$
%\item
%\emph{UNDEFINED}: The point does not have a value. %The possibility distribution for a given fuzzy number $X$ is $\pi_X=0$
%\end{itemize}
%

@@

\begin{table}
\centering
\caption{Relational representation for an ill-known time point. FT denotes Fuzzy Type. Field FT indicates that the values stored in F1, F2 and F3 denote either NULL, UNKNOWN, UNDEFINED or a triangular possibility distribution $M$ = $\left[D,\ a,\ b \right]$. Fields F1, F2 and F3 contain the actual values.}%  A \emph{PVP} is represented by two ill-known points.}
\begin{tabular}{c c c c c c l p{2cm}}
\hline
Value & FT & F1 & F2 & F3 & $\mu(x)$ & Description \\ \hline
UNKNOWN & 0 & N & N & N  & $1$ & Any value is equally possible\\ 
UNDEFINED & 1 & N & N & N & $0$ & The value is not defined \\%in the \\
    %      &   &   &   &   &     & attribute domain\\ 
NULL & 2 & N & N & N &not defined & Nothing is known about the value \\ 
$M$ & 3 & $D$ & $a$ & $b$ & $\mu_{M}$ & Ill-known value \\ 
\hline 
\end{tabular}
\label{table:relational-representation-pvp}

@@

\subsection{Querying Ill-known Valid-time Intervals}
This subsection discusses a tool for querying. Although TSQL~\cite{Jensen1995} is the temporal extension to SQL, the query language proposed is an extension to FSQL~\cite{Gal98} with temporal operators~\cite{Garrido2009}. The focus will evidently lie on the querying of valid-time periods. First, the query structure will be defined, followed by the evaluation and ranking methods.

%In this subsection we will define the query specification, then the evaluation of the query and finally the ranking for the query.

%It is important to notice that every Valid-time notion in the database is represented using a PVP. Thus, Valid-time intervals in the database can be partially unknown.

%In the query, the user is allowed to specify both non-temporal preferences and a temporal constraint.

%This allows the user to specify both the preferences and an ill-known valid-time interval in the query. It is important to notice that the possibilistic / fuzzy data stored in the database has a \emph{disjunctive interpretation} (it is said that we have \emph{uncertainty}: the valid-time interval has only one value but, for some reason the value is ill-known). In the query specification, the user is allowed to express a crisp time interval.

@@

% one is a temporal constraint and the other is the collection of query specifications for regular attributes:

%the first one is the temporal specification. The second part is the query specification for regular attributes.

\begin{definition}
A query $\tilde Q$ is specified by:
\begin{equation}
\label{eq:query-definition}
\tilde Q = \left( Q^{time}, Q \right)
\end{equation}
\end{definition}
Here, $Q$ denotes the collection of (possibly fuzzy) non-temporal preferences of the user and $Q^{time}$ denotes the temporal constraint given by the user:
\begin{definition}
 $Q^{time}$ is defined by:
\begin{equation}
Q^{time} = \left( I , AR \right)
\end{equation}
\end{definition}
Here, $I$ is a crisp time interval and $AR$ is one of Allen's relations \cite{Allen83}. The interpretation is that for a record with valid-time period given by a PVP $J$, the user requires that $I$ $AR$ $J$ holds.

%In table \ref{tab:allen-relations}, .

%\vspace{-35pt}

\subsubsection{Query evaluation}
In fuzzy querying of regular (relational) databases, the modelling of query satisfaction is a matter of degree. Usually, the evaluation of the query requirements for a record results in a satisfaction degree $s$, where $s$ lies in $\left[0,1\right]$, where 0 denotes total dissatisfaction and 1 denotes complete satisfaction. In crisp querying, the evaluation of query requirements for a record results in the accepting or rejecting of the record as a part of the result set. This can be modelled using satisfaction degrees, by assigning rejection a degree of $0$ and acceptance a degree of $1$ and not using any other value in $\left[0,1\right]$.

@@

% by a combination of non-temporal query requirements $Q$ is noted $e_{Q}(r) \in \left[0,1\right]$ in this work. The total query evaluation method is now the following.  and Allen relation $AR$
\begin{itemize}
\item
The preferences expressed in $Q$ are evaluated, resulting in a satisfaction degree denoted here $e_{Q}(r)$. The presented model accepts any sound way of calculating this evaluation, as long as $e_{Q}(r) \in \left[0,1\right]$. 
\item
Depending on AR, a specific set of ill-known constraints is considered, which can be found in Table 2. The possibility and necessity that $r$ fulfills all these constraints are calculated using formulas based on equations \eqref{ill-known-pos} respectively \eqref{ill-known-nec} and aggregated using the $\min$ operator.

%The sum of the possibility score and the necessity score gives an evaluation score $e_{Q^{time}}(r)$ in interval $\left[0,2\right]$. Because necessity cannot exceed $0$ unless possibility is $1$, this gives a natural ranking score.
\end{itemize}

@@

\subsubsection{Ranking and aggregation}
In order to present the results to the user, a crude ranking method is used: for every record $r$, the sum of $\Pos_{Q^{time}}(r)$ and $\Nec_{Q^{time}}(r)$ gives an evaluation score $e'_{Q^{time}}(r)$ in interval $\left[0,2\right]$. Because necessity cannot exceed $0$ unless possibility is $1$, this gives a natural ranking score. Some authors ~\cite{Bosc2010a} mentioned before that the possibility and necessity measures result in a total order in the set of events.  This $e'_{Q^{time}}(r)$ is then rescaled to the unit interval, resulting in $e_{Q^{time}}(r)$. The final ranking $e_{final}(r)$ is now given by a convex combination:

%the possibility and the necessity are ranked with the convex combination in \eqref{eq:convex-comb} with $\omega=0.5$. Finally this value is aggregated with the rest of the criteria with the same combination. %In this case the value for the aggregation is also $\omega = 0.5$

@@

%By increasing the value for the parameter $\omega$, the preferences expressed in the query $Q$ can be given more importance. By lowering the value for $\omega$, the temporal criteria is emphasized.
%The following example illustrates the querying process.

%\subsection{Example}

\begin{example}

%\subsubsection{The database}
%\paragraph{The database}
Consider the example relation given in Table \ref{tb:car-models} describing car models, containing general attributes (model name, manufacturer, car segment) and one temporal attribute (stored as explained in subsection \ref{subsec:storage}) describing the approximate period of time during which the car model was sold. The value for $D$ is stored in \emph{yyyy} format and $a$ and $b$ are represented by an integer. The ID field identifies a car model while the field Instance ID (IID) identifies the instance for a car model, thus a car model in a certain state.

\end{example}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Sample table for the car models. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
\centering
\caption{Example database}
\begin{tabular}{c c c c c c c}
\hline
ID & IID & Segment & Manufacturer & Name & Start & End  \\ [0.5ex]
\hline
001 & 1 & B & Peugeot & 205 & [1985,2,3] & [1997,2,1] \\
002 & 1 & C & Peugeot & 305 & [1977,2,2] & [1989,2,3] \\
003 & 1 & B & Citroen & C2 & [2002,2,2] & [2006,1,1] \\
001 & 2 & B & Peugeot & 206 & [2000,1,2] & [2011,2,1] \\
001 & 3 & B & Peugeot & 207 & [2006,1,1] & [2011,1,1]\\
\hline
\end{tabular}
\vspace{10pt}
\label{tb:car-models}
\vspace{-30pt}
\end{table}

%\subsubsection{The query}
%\paragraph{The query}
Consider the following query:
\begin{center}
\emph{The user wants to obtain a list of models from segment B, made by manufacturer Peugeot before the year interval 2001-2005}\\
\end{center}

Using the introduced notations in \eqref{eq:query-definition}, the query is translated to: %the following expression:

%\vspace{-10pt}
%
%\begin{align}
%$\tilde{Q} = \left(Q^{time}, Q\right)$
%\end{center}
%
%\vspace{-10pt}
%
%In this expression:
%\begin{itemize}
%\item
%$Q^{time}\ = \ ( \left[ 2001,\ 2005 \right],\ $  before $)$.
%\item
%$c_{segment}\ = \ $ Peugeot.
%\end{itemize}

%\vspace{-15pt}

@@

%The evaluation for the criteria are presented in the result table \ref{tb:results}.

@@

Table \ref{tb:results} shows a natural and gradual ranking for the results. The last record, (ID 001, IID 3) shows also that with $\omega$ = 0.5 both temporal and regular criteria have the same importance.% By lowering the value for omega the temporal criteria would have more impact on the final ranking.

>>>>

In this section, a valid-time model, able to handle imperfect valid-time notions, will be presented. First, the representation and storage of valid-time will be presented. Next, an approach to query the model is proposed. Finally, an example is given.

@@

As mentioned in \cite{Pons2011}, this approach differs from the one where a valid-time period is represented by one fuzzy set. Such a fuzzy set is seen as a possibility distribution on $\mathbb{R}$ and thus defines just one ill-known value. However, in the presented approach, a time period is modelled using an ill-known set, which is defined by a possibility distribution on $\Pow(\mathbb{R})$.

\subsubsection{Storage}
\label{subsec:storage}
To store a PVP, the possibility distributions defining the ill-known start and end point are stored. To store such a possibility distribution, the representation as used in the fuzzy interface for relational databases \emph{FIRST}~\cite{Medina1994}, \cite{Galindo1998} is used. Using this representation, it is possible to represent not only fuzzy numbers, but also (fuzzy) constants. To store an ill-known value, four values (FT, F1, F2 and F3) are stored. They are explained in Table \ref{table:relational-representation-pvp}. Note that while NULL denotes the fuzzy constant NULL, N denotes a null value in the database \cite{Medina1994}.

@@

\begin{table}
\centering
\caption{Relational representation for an ill-known time point. FT denotes Fuzzy Type. Field FT indicates that the values stored in F1, F2 and F3 denote either NULL, UNKNOWN, UNDEFINED or a triangular possibility distribution $M$ = $\left[D,\ a,\ b \right]$. Fields F1, F2 and F3 contain the actual values.}

\begin{tabular}{c c c c c c l p{2cm}}
\hline
Value & FT & F1 & F2 & F3 & $\mu(x)$ & Description \\ \hline
UNKNOWN & 0 & N & N & N  & $1$ & Any value is equally possible\\ 
UNDEFINED & 1 & N & N & N & $0$ & The value is not defined \\
NULL & 2 & N & N & N &not defined & Nothing is known about the value \\ 
$M$ & 3 & $D$ & $a$ & $b$ & $\mu_{M}$ & Ill-known value \\ 
\hline 
\end{tabular}
\label{table:relational-representation-pvp}

@@

\subsection{Querying Ill-known Valid-time Intervals}
This subsection discusses a tool for querying. The focus will evidently lie on the querying of valid-time periods. First, the query structure will be defined, followed by the evaluation and ranking methods.

@@

\begin{definition}
A query $\tilde Q$ is specified by:
\begin{equation}
\label{eq:query-definition}
\tilde Q = \left( Q^{time}, Q \right)
\end{equation}
\end{definition}
Here, $Q$ denotes the collection of (possibly fuzzy) non-temporal preferences of the user and $Q^{time}$ denotes the temporal constraint given by the user:
\begin{definition}
 $Q^{time}$ is defined by:
\begin{equation}
Q^{time} = \left( I , AR \right)
\end{equation}
\end{definition}
Here, $I$ is a crisp time interval and $AR$ is one of Allen's relations \cite{Allen1983}. The interpretation is that for a record with valid-time period given by a PVP $J$, the user requires that $I$ $AR$ $J$ holds.

\subsubsection{Query Evaluation}
In fuzzy querying of regular (relational) databases, the modelling of query satisfaction is a matter of degree. Usually, the evaluation of the query requirements for a record results in a satisfaction degree $s$, where $s$ lies in $\left[0,1\right]$, where 0 denotes total dissatisfaction and 1 denotes complete satisfaction. In crisp querying, the evaluation of query requirements for a record results in the accepting or rejecting of the record as a part of the result set. This can be modelled using satisfaction degrees, by assigning rejection a degree of $0$ and acceptance a degree of $1$ and not using any other value in $\left[0,1\right]$.

@@

\begin{itemize}
\item
The preferences expressed in $Q$ are evaluated, resulting in a satisfaction degree denoted here $e_{Q}(r)$. The presented model accepts any sound way of calculating this evaluation, as long as $e_{Q}(r) \in \left[0,1\right]$. 
\item
Depending on AR, a specific set of ill-known constraints is considered, which can be found in Table 2. The possibility and necessity that $r$ fulfills all these constraints are calculated using formulas based on equations \eqref{ill-known-pos} respectively \eqref{ill-known-nec} and aggregated using the $\min$ operator. 
\end{itemize}

@@

\subsubsection{Ranking and Aggregation}
In order to present the results to the user, a crude ranking method is used: for every record $r$, the sum of $\Pos_{Q^{time}}(r)$ and $\Nec_{Q^{time}}(r)$ gives an evaluation score $e'_{Q^{time}}(r)$ in interval $\left[0,2\right]$. Because necessity cannot exceed $0$ unless possibility is $1$, this gives a natural ranking score. Some authors ~\cite{Bosc2010a} mentioned before that the possibility and necessity measures result in a total order in the set of events.  This $e'_{Q^{time}}(r)$ is then rescaled to the unit interval, resulting in $e_{Q^{time}}(r)$. The final ranking $e_{final}(r)$ is now given by a convex combination:

@@

\begin{example} 
Consider the example relation given in Table \ref{tb:car-models} describing car models, containing general attributes (model name, manufacturer, car segment) and one temporal attribute (stored as explained in subsection \ref{subsec:storage}) describing the approximate period of time during which the car model was sold. The value for $D$ is stored in \emph{yyyy} format and $a$ and $b$ are represented by an integer. The ID field identifies a car model while the field Instance ID (IID) identifies the instance for a car model, thus a car model in a certain state.
\end{example}

\begin{table}[h]
\centering
\caption{Example database}
\begin{tabular}{c c c c c c c}
\hline
ID & IID & Segment & Manufacturer & Name & Start & End  \\ [0.5ex]
\hline
001 & 1 & B & Peugeot & 205 & [1985,2,3] & [1997,2,1] \\
002 & 1 & C & Peugeot & 305 & [1977,2,2] & [1989,2,3] \\
003 & 1 & B & Citroen & C2 & [2002,2,2] & [2006,1,1] \\
001 & 2 & B & Peugeot & 206 & [2000,1,2] & [2011,2,1] \\
001 & 3 & B & Peugeot & 207 & [2006,1,1] & [2011,1,1]\\
\hline
\end{tabular}
\vspace{10pt}
\label{tb:car-models}
\vspace{-30pt}
\end{table}

Consider the following query:
\begin{center}
\emph{The user wants to obtain a list of models from segment B, made by manufacturer Peugeot before the year interval 2001-2005}\\
\end{center}

Using the introduced notations in \eqref{eq:query-definition}, the query is translated to:

@@

@@

Table \ref{tb:results} shows a natural and gradual ranking for the results. The last record, (ID 001, IID 3) shows also that with $\omega$ = 0.5 both temporal and regular criteria have the same importance.
